Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2018-09-07 paper/tidy-missing-data-paper.Rmd
minor changes from submission to JCGS - mostly typos / changing some devices to PNG

---
title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
date: "September 6, 2018"
toc: false
bibliography: tidy-missing-data-paper.bib
output:
  bookdown::pdf_document2:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
csl: jcgs.csl
---

**Abstract**:

Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.

_Keywords:_ workflow, statistical computing, data science, data visualization, tidyverse, data pipeline

@@

`ggplot2` incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, `ggplot2` omits the missing values and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this.

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses visualizations that are easy to remember and compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews \@ref(overview-plots); univariate \@ref(univariate-plots), bivariate \@ref(bivariate-plots), and multivariate \@ref(multivariate-plots). All plots are created using `ggplot2`, giving users clearer control over the plot appearance, and customizability.

@@

There are different ways to summarize the amount of missing and complete values in a dataset. This section describes approaches to summarizing missingness, implemented in the `naniar` package. The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe. This means they integrate well with plotting and modelling tools. Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.

@@

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figure \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).   
```r
ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")

>>>>

---
title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
author:
- familyname: Tierney
  othernames: Nicholas
  address: Monash University
  email: nicholas.tierney@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University
abstract: "Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`."
keywords: "workflow, statistical computing, data science, data visualization, tidyverse, data pipeline"
wpnumber: 1/2018
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: tidy-missing-data-paper.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

`ggplot2` incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, `ggplot2` omits the missing values and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this.

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses visualizations that are easy to remember and compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)). All plots are created using `ggplot2`, giving users clearer control over the plot appearance, and customizability.

@@

There are different ways to summarize the amount of missing and complete values in a dataset. This section describes approaches to summarizing missingness, implemented in the `naniar` package. The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe. This means they integrate well with plotting and modelling tools. Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.

@@

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).   
```r
ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")
